% !TEX TS-program = pdflatex
% !TEX encoding = UTF-8 Unicode
% !BIB TS-program = bibtex
%
% ============================================================================
% paper.tex -- A universal spectral lower bound on graph-supervised learning
%
% Target  : Journal of Machine Learning Research (JMLR) or Foundations
%           of Computational Mathematics (FoCM).
% Authors : Rohan Fossé, Gaël Pallares, CESI LINEACT, 2026.
%
% Source notes (absorbed):
%   notes/01_universal_bound_proof.tex          -- Theorem 1 (universal LB)
%   notes/02_minimax_saturation.tex             -- Theorem 2 (corrected UB),
%                                                  Theorem 3 (Berry-Esseen LB)
%   notes/03_non_realisable_resolution.tex      -- Vacuous non-realisable
%                                                  regime; learner-specific
%                                                  formulation
% ============================================================================

\documentclass[11pt,a4paper]{article}

\usepackage[margin=2.5cm]{geometry}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{microtype}
\usepackage{amssymb,amsmath,amstext,amsthm,mathtools}
\usepackage{bm}
\usepackage{graphicx}
\usepackage{booktabs,array,float}
\usepackage{enumitem}
\setlist{topsep=2pt,itemsep=2pt,parsep=0pt}
\usepackage{hyperref}
\usepackage{xcolor}
\hypersetup{colorlinks=true,linkcolor=blue!50!black,citecolor=blue!50!black,urlcolor=blue!50!black}

\usepackage[authoryear,round]{natbib}
\bibliographystyle{plainnat}

\graphicspath{{figures/}}

% Theorem environments
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{conjecture}[theorem]{Conjecture}

% Notation macros (mirror notes/0X and the MLST companion paper)
\newcommand{\bphi}{\boldsymbol{\phi}}
\newcommand{\by}{\mathbf{y}}
\newcommand{\bP}{\mathbf{P}}
\newcommand{\bB}{\mathbf{B}}
\newcommand{\bU}{\mathbf{U}}
\newcommand{\bI}{\mathbf{I}}
\newcommand{\bM}{\mathbf{M}}
\newcommand{\bv}{\mathbf{v}}
\newcommand{\bu}{\mathbf{u}}
\newcommand{\bc}{\mathbf{c}}
\newcommand{\bX}{\mathbf{X}}
\newcommand{\bD}{\mathbf{D}}
\newcommand{\bW}{\mathbf{W}}
\newcommand{\bone}{\mathbf{1}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\PR}{\mathbb{P}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\Cov}{\mathrm{Cov}}
\newcommand{\tr}{\mathrm{tr}}
\newcommand{\Rspec}{R^2_{\mathrm{spec}}}
\newcommand{\Lsym}{\mathcal{L}_{\mathrm{sym}}}
\newcommand{\Scal}{\mathcal{S}}
\newcommand{\Acal}{\mathcal{A}}
\newcommand{\Hcal}{\mathcal{H}}
\newcommand{\Tcal}{\mathcal{T}}
\newcommand{\Dcal}{\mathcal{D}}
\newcommand{\Ycal}{\mathcal{Y}}
\newcommand{\Rcal}{\mathcal{R}}
\newcommand{\frR}{\mathfrak{R}}

\title{A universal spectral lower bound on the leave-node-out\\
generalisation error of graph-supervised learning}

\author{Rohan Foss\'e\thanks{rfosse@cesi.fr; ORCID 0009-0002-2195-0198.}
\and Ga\"el Pallares\thanks{ORCID 0009-0002-8680-604X.}\\
\small CESI LINEACT (EA 7527), Montpellier, France}

\date{Working draft v0.1 --- May 2026}

\begin{document}

\maketitle

\begin{abstract}
We prove a universal lower bound on the expected leave-node-out
generalisation error of any feature-restricted learner on any
graph-supervised regression problem.  The bound depends only on three
objects: the graph Laplacian, the target signal, and the learner's
reachable feature subspace.  It is independent of the regressor choice
and exact in expectation, with no concentration slack.  We prove that
an empirical-risk-minimising learner saturates the bound to a
Berry--Esseen slack of order $\rho(\Pi) \cdot M^2/\sqrt{n}$, and that
this slack is itself minimax-tight: even an oracle estimator with full
knowledge of the target on the population cannot drive it lower.
The pair of bounds is the Cram\'er--Rao analog, for graph-supervised
learning under leave-node-out evaluation, of the classical efficiency
inequality for parameter estimation.  Four corollaries instantiate the
framework: the encoding-gap bound for materials-property prediction
(recovering~\citealp{FossePallares2026applicabilityBound}), a spectral-disjointness
bound on negative transfer, a label-complexity dual for active learning,
and an empirical validation across the eight-task MatBench panel,
27-network bike-share demand panel, and 34{,}858-commune French mobility
panel established by the companion repositories of the cycling-data-lab
organisation.
\end{abstract}

\noindent\textbf{Keywords:} graph-supervised learning; generalisation
lower bound; leave-node-out cross-validation; spectral methods; Laplacian
eigenfunctions; Berry--Esseen anticoncentration; Cram\'er--Rao analog;
applicability domain; transfer learning; active learning.

% =========================================================================
\section{Introduction}
\label{sec:intro}

Generalisation bounds for supervised learning come, with rare exceptions,
in two methodological lanes.  The VC / Rademacher lane
\citep{Vapnik1998,Mohri2018book,Tripuraneni2020} bounds the gap between
empirical and population risk in terms of the hypothesis class's
combinatorial or geometric complexity, yielding \emph{upper bounds} on
the excess risk.  The information-theoretic lane
\citep{Russo2016,Xu2017,Tran2019,Nguyen2023} bounds the same gap in terms
of the mutual information between the training data and the learned
hypothesis, again yielding upper bounds.  In both cases, the resulting
inequalities tell us how \emph{badly} a learner could perform; they do
not characterise an irreducible floor below which no learner can go.

In a parallel development at the intersection of graph signal processing
\citep{Shuman2013,Ortega2018} and statistical learning, the question of
\emph{leave-node-out generalisation} on a graph-supported regression
problem has acquired particular significance.  The materials-informatics
community has investigated the \emph{applicability-domain problem}
\citep{Tropsha2010,Sahigara2012,Hanser2016} --- the empirical observation
that models trained on one family of compositions generalise poorly to
materials outside that family --- through largely empirical means.  The
recommender-systems community has investigated the \emph{cold-start
problem} \citep{Schein2002,Marlin2003,Kluver2014} as a structurally
similar question on user--item graphs.  Urban mobility research has
investigated the failure of mode-choice models trained in one city to
generalise to others \citep{Naik2014,Hillel2021}.  In every case the
empirical phenomenology is unambiguous, but a quantitative theorem
characterising the \emph{floor} of the held-out error --- as a function
of the data, not of the regressor --- was, until recently, missing.

The companion paper \citet{FossePallares2026applicabilityBound} provided one
such floor, specifically for the materials-informatics setting.  The
present paper generalises that result to a universal statement that
applies to any graph-supervised regression problem under leave-node-out
evaluation, and shows that the saturating estimator is in principle any
empirical-risk minimiser whose hypothesis class subsumes the population-risk
minimiser on the relevant feature subspace.  Together with a matching
Berry--Esseen minimax lower bound, this yields a Cram\'er--Rao-style
efficiency statement for graph-supervised learning.

\subsection{Contributions}

\begin{enumerate}[label=(\arabic*),leftmargin=*]
\item \textbf{Universal lower bound
    (Theorem~\ref{thm:universal-lower-bound}).}  For any learner $\Acal$
    with reachable hypothesis class $\Hcal_d$, target $\by \in \R^N$
    with $\|\by\|_\infty \leq M$, and uniform leave-node-out protocol
    $\Pi_{\mathrm{LSO}}$ with $|T| = n$, the expected leave-node-out loss
    satisfies
    \begin{equation*}
        \E_T\!\big[L_{T^c}(\hat f_\Acal(T))\big] \;\geq\; \big(1 - \Rspec(\Hcal_d, \by)\big) \cdot \Var(\by),
    \end{equation*}
    where $\Rspec(\Hcal_d, \by) := 1 - L_V(f^\star_{\Hcal_d})/\Var(\by)$
    and $f^\star_{\Hcal_d} := \arg\min_{f \in \Hcal_d} L_V(f)$.  The
    bound is \emph{exact in expectation}: no concentration slack
    appears.  Proved as
    Theorem~\ref{thm:universal-lower-bound} in
    Section~\ref{sec:lower-bound-proof}.

\item \textbf{Universal upper bound
    (Theorem~\ref{thm:universal-upper-bound}).}  Any empirical-risk-minimising
    learner with truncated $L$-Lipschitz hypothesis class $\Hcal_d$ of
    Rademacher complexity $\mathcal{R}_n(\Hcal_d)$ satisfies, with
    probability $\geq 1 - \eta$ over $T \sim \Pi_{\mathrm{LSO}}$,
    \begin{equation*}
        L_{T^c}(\hat f_\Acal(T))
        \;\leq\; \big(1 - \Rspec(\Hcal_d, \by)\big) \cdot \Var(\by)
            + \rho(\Pi) \cdot \!\left[ 2 \mathcal{R}_n(\Hcal_d) + 3 M^2 \sqrt{\tfrac{\log(2/\eta)}{2 n}} \right] + O\!\left(\tfrac{M^2}{\sqrt{n}}\right),
    \end{equation*}
    where $\rho(\Pi) := N/(N-n)$ is the transduction factor.  Proved
    in Section~\ref{sec:upper-bound-proof}.

\item \textbf{Minimax tightness (Theorem~\ref{thm:minimax-lower-bound}).}
    No estimator --- including an oracle with full population knowledge
    of $\by$ --- can drive the slack of Theorem~\ref{thm:universal-upper-bound}
    below $\Omega(M^2/\sqrt{N-n})$ in the worst case over signals.  The
    argument is a Berry--Esseen anticoncentration on the residual energy
    $\|\by^\star_\perp\|^2_{T^c}$, the fluctuation of which is intrinsic
    to the random test set, not estimator-dependent.  Proved in
    Section~\ref{sec:minimax-proof}.

\item \textbf{Cram\'er--Rao analogy.}  Theorems
    \ref{thm:universal-lower-bound}--\ref{thm:minimax-lower-bound} are
    the graph-supervised-learning analog of the classical efficiency
    inequality $\Var(\hat\theta) \geq 1/I(\theta)$:
    Theorem~\ref{thm:universal-lower-bound} provides a structural
    floor on the expected loss depending only on (graph, target,
    learner's reach), and Theorem~\ref{thm:universal-upper-bound}
    exhibits an explicit efficient estimator achieving the floor up to a
    minimax-tight slack.  The structural role of the Fisher information
    is played by the variance of the residual energy
    $((\bP_{\Hcal_d^\perp}\by)(v))^2$ on $V$.

\item \textbf{Corollaries.}
    \begin{itemize}[label=$\bullet$,leftmargin=*]
    \item \textbf{C1 (encoding-gap recovery; \S\ref{subsec:c1}).}  The
        bound of \citet{FossePallares2026applicabilityBound} on the categorical
        vs.\ compositional encoding gap under leave-composition-out
        evaluation is a one-paragraph corollary, with the categorical
        learner the degenerate case $\Hcal_d = \{$constants$\}$ and
        the compositional learner the case $\Hcal_d = \Scal_{\mathrm{comp}}$.
    \item \textbf{C2 (negative transfer; \S\ref{subsec:c2}).}  Taking
        $V = V_{\mathrm{src}} \cup V_{\mathrm{tgt}}$ and $\Pi$ the
        leave-task-out protocol, the bound recovers a structural
        \emph{spectral-disjointness} criterion for negative transfer:
        transfer is unavoidably harmful when the source and target
        labels project onto disjoint bands of the joint graph's
        Laplacian eigenbasis.
    \item \textbf{C3 (active learning; \S\ref{subsec:c3}).}  Letting the
        protocol $\Pi$ be chosen by the learner gives an
        importance-weighted variant in which the optimal sampling
        distribution is the leverage-score distribution on the learner's
        reachable subspace, with the bound matching the rate of the
        spectral-graph A-optimal design literature
        \citep{DrineasMahoney2018,Avron2017}.
    \end{itemize}

\item \textbf{Empirical validation.}  The companion repositories
    \texttt{materials-applicability-bound}
    \citep{FossePallares2026applicabilityBound} and
    \texttt{mobility-applicability-bound} (Foss\'e and Pallares, in
    preparation) provide cross-domain empirical instantiations on the
    8-task MatBench panel, 27-network bike-share demand panel, and
    34{,}858-commune French mobility panel respectively.
    Section~\ref{sec:empirics} summarises the headline results and
    points to the per-repository scripts for reproduction.
\end{enumerate}

\subsection{Relation to prior work}
\label{subsec:related-work}

The four nearest methodological lanes are:

\textbf{(a) Rademacher / VC bounds.}  Classical generalisation bounds
\citep{Vapnik1998,Mohri2018book} and the modern transfer-learning
extensions \citep{Tripuraneni2020,Du2020} give \emph{upper} bounds on
excess risk via Rademacher complexity, eluder dimension, or task
diversity.  Our Theorem~\ref{thm:universal-upper-bound} reuses the
Rademacher technology to prove its upper bound, but the principal
contribution is the matching lower bound and the resulting
Cram\'er--Rao-style efficiency statement, which has no analog in the
VC lane.

\textbf{(b) Information-theoretic bounds.}  Recent
information-theoretic bounds on transferability
\citep{Tran2019,Nguyen2023} and on generalisation
\citep{Russo2016,Xu2017} provide model-free upper bounds via conditional
entropy or PAC-Bayes.  Our framework is also model-free, but in the
spectral lane rather than the information-theoretic lane, and provides
lower rather than upper bounds.

\textbf{(c) Graph signal processing on bandlimited subspaces.}  The
classical bandlimited reconstruction theorems of
\citet{Pesenson2008,AnisGaddeOrtega2016} and the recent generalisation
to arbitrary feature subspaces by \citet{ChepuriLeus2018,TanakaEldar2020}
provide the underlying sampling-theoretic machinery for graph signals
restricted to a subspace.  Our framework borrows the notion of a
subspace-restricted estimator but, crucially, does \emph{not} require
the Pesenson sampling quality $\sigma_T(\Hcal_d)$ to appear in the
bound; see Remark~\ref{rem:no-sigma-T-needed} for the resolution of an
intermediate misconception in our own development.

\textbf{(d) Cycling-data-lab structural-bounds program.}  This paper is
the foundational entry of the cycling-data-lab structural-bounds
research program.  It absorbs and unifies the empirical
applicability-domain bound of \citet{FossePallares2026applicabilityBound} (MLST,
materials informatics) and the mobility-transfer instantiation in
preparation by the same authors (TR-B, urban mobility).  The bound
established here is the \emph{floor}; the empirical companions
demonstrate that the floor is operationally tight in two unrelated
domains.

\subsection{Roadmap}

Section~\ref{sec:setup} fixes notation and assumptions.
Section~\ref{sec:main-results} states the three main theorems and the
Cram\'er--Rao corollary.  Sections~\ref{sec:lower-bound-proof},
\ref{sec:upper-bound-proof}, \ref{sec:minimax-proof} contain the proofs.
Section~\ref{sec:corollaries} develops the three corollaries C1--C3.
Section~\ref{sec:empirics} summarises empirical validation.
Section~\ref{sec:discussion} discusses open questions and limitations.

% =========================================================================
\section{Setup}
\label{sec:setup}

\subsection{Graph, signal, and spectrum}

Let $G = (V, E)$ be a finite undirected graph with $|V| = N$.  We treat
$V$ as the set of supervised examples and $G$ as a similarity graph
chosen by the modeller: in materials informatics, a composition-similarity
$k$-NN graph; in urban mobility, a commune-similarity graph; in
recommender systems, a user--user (or item--item) similarity graph.

The symmetric-normalised Laplacian is
$\Lsym = \bI - \bD^{-1/2} \bW \bD^{-1/2}$, where $\bW$ is the
weighted adjacency matrix and $\bD$ the diagonal degree matrix.  Its
eigendecomposition is $\Lsym = \bU \boldsymbol{\Lambda} \bU^\top$ with
$0 = \lambda_1 \leq \lambda_2 \leq \cdots \leq \lambda_N \leq 2$ and
$\bU = [\bu_1, \ldots, \bu_N]$ orthonormal.

The target signal is a vector $\by \in \R^N$, treated as deterministic
(not random).  Without loss of generality we assume $\by$ is centred,
$\bone^\top \by = 0$ (this absorbs any constant predictor into the
reachable subspace; see Remark~\ref{rem:centring}).  Under this
convention, the population variance
$\Var(\by) := N^{-1} \|\by\|^2$.
We assume the signal is bounded: $\|\by\|_\infty \leq M < \infty$.

The graph Fourier coefficients of $\by$ are $\hat y_k := \bu_k^\top \by$,
so $\by = \sum_k \hat y_k \bu_k$ and $\|\by\|^2 = \sum_k \hat y_k^2$
by Parseval.

\subsection{Evaluation protocol}
\label{subsec:setup-protocol}

\begin{definition}[Evaluation protocol]
\label{def:protocol}
An \emph{evaluation protocol} is a distribution $\Dcal_\Pi$ over
ordered pairs $(T, T^c)$ with $T \subset V$, $|T| = n$,
$T^c := V \setminus T$, $|T^c| = N - n$.  The canonical instance
$\Pi_{\mathrm{LSO}}$ samples $T$ uniformly without replacement.
\end{definition}

For a predictor $\hat f : V \to \R$ (identified with its vector of
node predictions in $\R^N$), we write
$L_T(\hat f) := n^{-1} \sum_{v \in T} (\hat f(v) - y_v)^2$,
$L_{T^c}(\hat f) := (N-n)^{-1} \sum_{v \in T^c} (\hat f(v) - y_v)^2$,
and $L_V(\hat f) := N^{-1} \sum_{v \in V} (\hat f(v) - y_v)^2$.
The \emph{transduction factor} is $\rho(\Pi) := N/(N-n)$; for the
standard 80/20 split, $\rho = 5$.

\subsection{Learner and reachable hypothesis class}
\label{subsec:setup-learner}

\begin{definition}[Learner]
\label{def:learner}
A \emph{learner} is a (possibly randomised) map $\Acal : (T, \by|_T)
\mapsto \hat f_\Acal(T) \in \Hcal_d$, where $\Hcal_d \subseteq \R^N$ is
the \emph{hypothesis class}: the set of vector-valued predictors the
algorithm can output, identified with their restrictions to $V$.  We
assume $\Hcal_d$ is closed under truncation to $[-M, M]$.
\end{definition}

\begin{definition}[Learner-specific spectral $R^2$]
\label{def:learner-rspec}
Let $f^\star_{\Hcal_d} := \arg\min_{f \in \Hcal_d} L_V(f)$ be the
\emph{population-risk minimiser} of $\Hcal_d$ (which exists by
compactness under truncation).  The \emph{learner-specific spectral
$R^2$} of $\Hcal_d$ on $\by$ is
\begin{equation}
    \Rspec(\Hcal_d, \by) \;:=\; 1 \,-\, \frac{L_V(f^\star_{\Hcal_d})}{\Var(\by)} \;\in\; [0, 1].
    \label{eq:rspec-learner-defn}
\end{equation}
For closed linear $\Hcal_d$, $f^\star_{\Hcal_d} = \bP_{\Hcal_d}\by$ and
$\Rspec(\Hcal_d, \by) = \|\bP_{\Hcal_d}\by\|^2/\|\by\|^2$, the classical
projection $R^2$.  For non-linear $\Hcal_d$ (gradient-boosted trees,
neural networks with weight constraints, $k$-NN with local averaging),
$\Rspec$ is the population-variance-explained ceiling of the class.
\end{definition}

\begin{remark}[Examples of $\Hcal_d$ and $\Rspec(\Hcal_d, \by)$]
\label{rem:examples}
\begin{itemize}[label=$\bullet$,leftmargin=*]
\item \emph{Categorical (one-hot) learner.}  $\Hcal_d = \{$constants on $T^c\}$
    (the learner sees one-hot indicators for $v \in T$ only, and is
    forced to a constant prediction on $T^c$).  Then
    $\Rspec(\Hcal_d, \by) = 0$ for centred $\by$.
\item \emph{Linear regressor on features $\bphi$.}
    $\Hcal_d = \{(\langle \bphi(v), \bm{\beta}\rangle)_{v \in V} : \bm{\beta} \in \R^d\}$;
    $\Rspec(\Hcal_d, \by)$ is the projection-$R^2$ of $\by$ on
    $\mathrm{col}(\bphi)$.
\item \emph{Kernel ridge with kernel $\mathcal{K}$.}  $\Hcal_d$ is the
    column span of the kernel matrix.
\item \emph{Sufficiently expressive MLP / GBM.}  $\Hcal_d = \R^N$, so
    $\Rspec(\Hcal_d, \by) = 1$ and the bound becomes vacuous --- as
    expected, since an arbitrarily expressive learner has no structural
    floor.
\end{itemize}
\end{remark}

\subsection{Assumptions}

\begin{assumption}[Bounded centred target]
\label{a:target}
$\|\by\|_\infty \leq M$ and $\bone^\top \by = 0$.
\end{assumption}

\begin{assumption}[ERM learner]
\label{a:erm}
The learner $\Acal$ is an empirical risk minimiser on $\Hcal_d$:
$\hat f_\Acal(T) \in \arg\min_{f \in \Hcal_d} L_T(f)$.  The hypothesis
class is truncated to $[-M, M]$ and admits a Rademacher complexity
$\mathcal{R}_n(\Hcal_d)$ in the sense of \citet[][\S 3.3]{Mohri2018book}.
\end{assumption}

% =========================================================================
\section{Main results}
\label{sec:main-results}

\begin{theorem}[Universal spectral lower bound]
\label{thm:universal-lower-bound}
Under Assumptions~\ref{a:target}--\ref{a:erm}, for any evaluation
protocol $\Pi$ satisfying $\E_T[L_T(f)] = L_V(f)$ for all deterministic
$f \in \R^N$ (in particular $\Pi_{\mathrm{LSO}}$),
\begin{equation}
    \E_{T \sim \Dcal_\Pi}\!\big[L_{T^c}(\hat f_\Acal(T))\big]
    \;\geq\; \big(1 - \Rspec(\Hcal_d, \by)\big) \cdot \Var(\by).
    \label{eq:universal-lower-bound}
\end{equation}
The bound is \emph{exact in expectation}: there is no concentration
slack term.
\end{theorem}

\begin{theorem}[Universal upper bound]
\label{thm:universal-upper-bound}
Under Assumptions~\ref{a:target}--\ref{a:erm}, with probability $\geq 1
- \eta$ over $T \sim \Pi_{\mathrm{LSO}}$,
\begin{equation}
    L_{T^c}(\hat f_\Acal(T))
    \;\leq\; \big(1 - \Rspec(\Hcal_d, \by)\big) \cdot \Var(\by)
        \;+\; \rho(\Pi) \cdot \Bigg[
            2 \mathcal{R}_n(\Hcal_d)
            + 3 M^2 \sqrt{\frac{\log(2/\eta)}{2 n}}
          \Bigg]
        + O\!\left(\frac{M^2}{\sqrt{n}}\right).
    \label{eq:universal-upper-bound}
\end{equation}
For learners satisfying $\mathcal{R}_n(\Hcal_d) = O(\sqrt{d/n})$
(linear, kernel-ridge, MLP with bounded weights, gradient-boosted
trees of bounded depth), the leading-order slack is
$O(\rho(\Pi) (M \sqrt{d} + M^2)/\sqrt{n})$.
\end{theorem}

\begin{theorem}[Berry--Esseen minimax lower bound on the slack]
\label{thm:minimax-lower-bound}
For any hypothesis class $\Hcal_d \subsetneq \R^N$, any $M > 0$, any
sample sizes $1 \leq n \leq N - 2$, and any confidence level $\eta \in
(0, 1/3)$, there exists a signal $\by^\star \in \Hcal_d^\perp$ with
$\|\by^\star\|_\infty = M$ such that
\begin{equation}
    \PR_T\!\Big[
        L_{T^c}(\hat f) - \big(1 - \Rspec(\Hcal_d, \by^\star)\big)\Var(\by^\star)
        \;\geq\; c \cdot M^2 / \sqrt{N - n}
    \Big]
    \;\geq\; 1 - \eta
    \label{eq:minimax-lower-bound}
\end{equation}
for an absolute constant $c > 0$ (Berry--Esseen, $c \approx 1/16$
suffices), uniformly in $\hat f$.  No estimator --- including one with
oracle access to $\by^\star$ on $V$ --- can drive the slack of
\eqref{eq:universal-upper-bound} below
$\Omega(M^2/\sqrt{N-n})$ in the worst case.
\end{theorem}

\begin{corollary}[Cram\'er--Rao efficiency for graph-supervised learning]
\label{cor:cramer-rao}
Combining Theorems~\ref{thm:universal-lower-bound}--\ref{thm:minimax-lower-bound},
the minimax slack rate at protocol $\Pi_{\mathrm{LSO}}$ on signal class
$\Ycal_{\Hcal_d, M} := \{\by \in \R^N : \|\by\|_\infty \leq M, \bone^\top
\by = 0\}$ is
\begin{equation}
    \frR_n(\Hcal_d, M, \eta)
    \;=\; \Theta\!\left(\frac{M^2}{\sqrt{N - n}} + \rho(\Pi) \mathcal{R}_n(\Hcal_d)\right),
    \label{eq:cramer-rao-rate}
\end{equation}
with absolute constants on both sides, matching up to the $\rho(\Pi)$
factor in the upper bound and the Berry--Esseen constant in the lower
bound.  The empirical-risk-minimising learner is therefore
minimax-efficient on graph-supervised regression problems under
leave-node-out, in the sense that the classical Cram\'er--Rao bound
characterises minimax-efficient estimators in regular parametric
estimation.
\end{corollary}

\begin{remark}[Centring]
\label{rem:centring}
If $\by$ is not centred, replace $\by$ throughout by $\by - \bar y \bone$,
where $\bar y := N^{-1}\bone^\top \by$.  Equivalently, augment
$\Hcal_d$ to contain the constant vector $\bone$, which absorbs the
mean predictor.  All standard learners satisfy this implicitly through
a bias / intercept term.
\end{remark}

\begin{remark}[Why the Pesenson sampling quality $\sigma_T$ does not appear]
\label{rem:no-sigma-T-needed}
An earlier formulation of Theorem~\ref{thm:universal-upper-bound}
(working notes \texttt{notes/01\_universal\_bound\_proof.tex} in the
companion repository, withdrawn in \texttt{notes/03\_non\_realisable\_resolution.tex})
expressed the slack as $O(\rho(\Pi) /
\sqrt{n \cdot \sigma_\Pi(\Hcal_d)})$, where $\sigma_\Pi(\Hcal_d) :=
\E_T[\lambda_{\min}(\bB^\top \bP_T \bB)]$ is the
Pesenson--Anis--Gadde--Ortega sampling quality of $\Pi$ for $\Hcal_d$.
This formulation was an artefact of using the Tikhonov-regularised
Pesenson-ridge estimator as the witness predictor.  In fact, the
population-risk minimiser $f^\star_{\Hcal_d}$ itself serves as a
witness without invoking the sampling-restricted inverse $(\bB^\top
\bP_T \bB)^{-1}$, eliminating the $\sigma_T$ factor entirely.  The
proof of Theorem~\ref{thm:universal-upper-bound} below uses this
cleaner witness, mirroring the proof of Theorem~1(c) of
\citet{FossePallares2026applicabilityBound}.  The
graph-Fisher-information / inverse-problem theory of
\citet{BauerPereverzev2005,Cavalier2008} --- relevant for noisy
continuous Hilbert-space inverse problems --- does not apply to the
finite-population empirical-risk problem treated here.
\end{remark}

% =========================================================================
\section{Proof of Theorem~\ref{thm:universal-lower-bound}}
\label{sec:lower-bound-proof}

The proof has three short steps.

\paragraph{Step 1 — Functional floor on $L_V$ (Bessel-projection inequality).}

\begin{lemma}[Bessel projection floor]
\label{lem:bessel}
For any $\hat f \in \Hcal_d$,
$N \cdot L_V(\hat f) = \|\hat f - \by\|^2 \;\geq\; \|f^\star_{\Hcal_d} - \by\|^2 = N \cdot (1 - \Rspec(\Hcal_d, \by))\Var(\by)$.
\end{lemma}

\begin{proof}
$f^\star_{\Hcal_d}$ is the population-risk minimiser by definition.
Hence $L_V(\hat f) \geq L_V(f^\star_{\Hcal_d}) = (1 - \Rspec(\Hcal_d,
\by))\Var(\by)$ for every $\hat f \in \Hcal_d$.  By Assumption~\ref{a:erm},
$\hat f_\Acal(T) \in \Hcal_d$, so the inequality applies pointwise in $T$.
\end{proof}

\paragraph{Step 2 — Transduction identity.}

\begin{lemma}[Population--train--holdout decomposition]
\label{lem:transduction}
For any deterministic $f \in \R^N$ and any $T \subset V$ with $|T| = n$,
\begin{equation}
    L_V(f) \;=\; \frac{n}{N}\, L_T(f) \;+\; \frac{N-n}{N}\, L_{T^c}(f),
    \label{eq:transduction-identity}
\end{equation}
equivalently $L_{T^c}(f) = \rho(\Pi) L_V(f) - (\rho(\Pi) - 1) L_T(f)$.
\end{lemma}

\begin{proof}
$N \cdot L_V(f) = \sum_{v \in V}(f(v) - y_v)^2 = \sum_{v \in T}(\ldots)
+ \sum_{v \in T^c}(\ldots) = n \cdot L_T(f) + (N-n) \cdot L_{T^c}(f)$.
\end{proof}

\paragraph{Step 3 — Combination.}

By Lemma~\ref{lem:transduction} applied to $f = \hat f_\Acal(T)$ and
taken in expectation over $T$,
\begin{equation}
    \E_T[L_{T^c}(\hat f_\Acal(T))]
    \;=\; \rho(\Pi) \cdot \E_T[L_V(\hat f_\Acal(T))]
        \;-\; (\rho(\Pi) - 1) \cdot \E_T[L_T(\hat f_\Acal(T))].
    \label{eq:lower-step3-expectation}
\end{equation}
By Lemma~\ref{lem:bessel}, $L_V(\hat f_\Acal(T)) \geq (1 -
\Rspec(\Hcal_d, \by))\Var(\by)$ pointwise in $T$, hence
$\E_T[L_V(\hat f_\Acal(T))] \geq (1 - \Rspec(\Hcal_d, \by))\Var(\by)$.

For the $L_T$ term, by Assumption~\ref{a:erm} (ERM property) and
$f^\star_{\Hcal_d} \in \Hcal_d$,
$L_T(\hat f_\Acal(T)) \leq L_T(f^\star_{\Hcal_d})$ pointwise in $T$.
Taking expectation and using the protocol hypothesis $\E_T[L_T(f)] = L_V(f)$
for deterministic $f$,
\begin{equation}
    \E_T[L_T(\hat f_\Acal(T))]
    \;\leq\; \E_T[L_T(f^\star_{\Hcal_d})]
    \;=\; L_V(f^\star_{\Hcal_d})
    \;=\; (1 - \Rspec(\Hcal_d, \by)) \Var(\by).
    \label{eq:lower-step3-LT}
\end{equation}
Setting $A := (1 - \Rspec(\Hcal_d, \by)) \Var(\by)$ and substituting
into \eqref{eq:lower-step3-expectation},
$$
\E_T[L_{T^c}(\hat f_\Acal(T))] \;\geq\; \rho(\Pi) \cdot A - (\rho(\Pi) - 1) \cdot A \;=\; A.
$$
This is Eq.~\eqref{eq:universal-lower-bound}.\qed

\begin{remark}[Why the bound is exact in expectation]
\label{rem:exact-in-expectation}
The two slack terms (from upper-bounding $L_V$ above by the Bessel
floor, and from upper-bounding $L_T$ below by the ERM-on-projection
inequality) cancel exactly through the transduction identity.  This
mirrors the structural unbiasedness condition under which the classical
Cram\'er--Rao bound holds with equality.  A high-probability version
of the bound carries a $O(\rho M^2/\sqrt{n})$ slack from the
Hoeffding--Serfling concentration of \eqref{eq:lower-step3-LT}; see
Theorem~\ref{thm:universal-upper-bound} for the matching upper bound.
\end{remark}

% =========================================================================
\section{Proof of Theorem~\ref{thm:universal-upper-bound}}
\label{sec:upper-bound-proof}

The proof structure mirrors that of part~(c) of Theorem~1 of
\citet[][\S 5.2]{FossePallares2026applicabilityBound}, with the
population-risk minimiser $f^\star_{\Hcal_d}$ replacing the specific
compositional projector $\bP_{\Scal_{\mathrm{comp}}}\by$ used there.

\paragraph{Step 1 — In-sample concentration of the witness.}

The population-risk minimiser $f^\star_{\Hcal_d} \in \Hcal_d$ is fixed
(not a function of $T$).  By Serfling's without-replacement Hoeffding
inequality \citep{Serfling1974,BardenetMaillard2015} applied to the
bounded sequence $\{(f^\star_{\Hcal_d}(v) - y_v)^2\}_{v \in V}$ ---
which has population mean $L_V(f^\star_{\Hcal_d}) = (1 -
\Rspec(\Hcal_d, \by))\Var(\by)$ and supremum at most $4 M^2$ --- with
probability $\geq 1 - \eta/2$ over $T \sim \Pi_{\mathrm{LSO}}$,
\begin{equation}
    L_T(f^\star_{\Hcal_d})
    \;\leq\; (1 - \Rspec(\Hcal_d, \by)) \Var(\by)
        + 4 M^2 \sqrt{\frac{(1 - (n-1)/N) \log(2/\eta)}{2 n}}.
    \label{eq:upper-step1-serfling}
\end{equation}

\paragraph{Step 2 — ERM property.}

By Assumption~\ref{a:erm}, $L_T(\hat f_\Acal(T)) \leq L_T(f^\star_{\Hcal_d})$
pointwise in $T$, since $f^\star_{\Hcal_d} \in \Hcal_d$ and ERM
minimises $L_T$ over $\Hcal_d$.

\paragraph{Step 3 — Rademacher generalisation.}

Mohri Theorem 3.7 \citep[][\S 3.3]{Mohri2018book} for the square
loss truncated to $[-M, M]^2$, combined with the Talagrand contraction
lemma \citep{Talagrand1996concentration} ($2M$-Lipschitz square loss),
gives with probability $\geq 1 - \eta/2$:
\begin{equation}
    L_V(\hat f_\Acal(T))
    \;\leq\; L_T(\hat f_\Acal(T))
        + 2 \mathcal{R}_n(\Hcal_d)
        + 3 M^2 \sqrt{\frac{\log(2/\eta)}{2 n}}.
    \label{eq:upper-step3-mohri}
\end{equation}

\paragraph{Step 4 — Transduction.}

By Lemma~\ref{lem:transduction}, $L_{T^c}(\hat f_\Acal(T)) = L_T(\hat
f_\Acal(T)) + \rho(\Pi)(L_V(\hat f_\Acal(T)) - L_T(\hat f_\Acal(T)))$.
Combining \eqref{eq:upper-step1-serfling}--\eqref{eq:upper-step3-mohri}
and a union bound across the two confidence events,
$$
L_{T^c}(\hat f_\Acal(T))
\;\leq\; (1 - \Rspec(\Hcal_d, \by))\Var(\by)
    + \rho(\Pi) \cdot \Big[ 2 \mathcal{R}_n(\Hcal_d)
      + 3 M^2 \sqrt{\log(2/\eta)/(2 n)} \Big]
    + O(M^2/\sqrt{n}),
$$
yielding \eqref{eq:universal-upper-bound}.\qed

% =========================================================================
\section{Proof of Theorem~\ref{thm:minimax-lower-bound}}
\label{sec:minimax-proof}

We construct an adversarial signal whose Berry--Esseen anticoncentration
on the held-out residual already forces an $\Omega(M^2/\sqrt{N-n})$
deviation between $L_{T^c}(\hat f)$ and its population value,
irrespective of estimator choice.

\paragraph{Construction of $\by^\star$.}

Pick any unit-infty vector $\bu \in \Hcal_d^\perp$ (exists because
$\Hcal_d \subsetneq \R^N$).  Rescale so $\bu$ has $N/2$ entries equal
to $\pm 1$ (in any sign pattern respecting $\bone^\top \bu = 0$) and
zero entries elsewhere.  Set $\by^\star := M \bu$.  Then $\by^\star
\in \Hcal_d^\perp$, $\|\by^\star\|_\infty = M$, $\bone^\top \by^\star
= 0$.  Since $\by^\star \in \Hcal_d^\perp$,
$\Rspec(\Hcal_d, \by^\star) = 0$ and the lower bound reference value is
$\Var(\by^\star) = M^2/2$.

\paragraph{Oracle slack.}

Any predictor $\hat f \in \Hcal_d$ (oracle or not) achieves population
loss $L_V(\hat f) \geq L_V(0) = \Var(\by^\star) = M^2/2$, since
$\by^\star \in \Hcal_d^\perp$ means the best in-class predictor is the
zero function.  Define $\Delta_T := L_{T^c}(0) - \Var(\by^\star)$.
For $\hat f = 0$,
$$
L_{T^c}(0) - \Var(\by^\star)
= \frac{1}{|T^c|}\sum_{v \in T^c} (y^\star_v)^2 - \frac{1}{N}\sum_v (y^\star_v)^2.
$$
Define $X_v := (y^\star_v)^2 - \Var(\by^\star) = M^2 \mathbf{1}_{\{v
\in \mathrm{supp}(\bu)\}} - M^2/2$.  The $X_v$ are centred ($\sum_v X_v
= 0$), bounded in $[-M^2/2, M^2/2]$, and have population variance
$N^{-1}\sum X_v^2 = M^4/4$.

\paragraph{Berry--Esseen on sampling without replacement.}

By the Berry--Esseen theorem for sampling without replacement
\citep{Bikelis1969,Bentkus2003}, the standardised sum
$\frac{1}{\sqrt{|T^c| M^4/4}}\sum_{v \in T^c} X_v$ converges to
$\mathcal{N}(0, 1)$ in Kolmogorov distance at rate $O(M^2/\sqrt{(N-n)
M^4/4}) = O(1/\sqrt{N-n})$.  For any threshold $\kappa$, the deviation
event
$$
\PR_T\!\left[\Delta_T \geq \kappa \cdot \frac{M^2}{2\sqrt{N-n}}\right]
$$
is, up to a Berry--Esseen correction of order $1/\sqrt{N-n}$, equal to
$\Phi(-\kappa)$.

\paragraph{Combine.}

For $\eta < 1/3$, choose $\kappa$ such that $\Phi(-\kappa) = 1 - \eta - 14/\sqrt{N-n}$
(absorbing the Berry--Esseen constant into the displayed constant).
For $N - n \geq 100$ and $\eta < 1/3$, $\kappa \geq 1$ suffices, giving
$$
\PR_T\!\left[\Delta_T \geq \frac{M^2}{2\sqrt{N-n}}\right] \;\geq\; 1 - \eta.
$$
Since the oracle predictor $\hat f = 0$ \emph{dominates} every other
$\hat f \in \Hcal_d$ on $\by^\star \in \Hcal_d^\perp$, this lower bound
applies to every learner, yielding \eqref{eq:minimax-lower-bound} with
$c = 1/2$ (resp.\ $c = 1/16$ uniformly across smaller $N - n$).\qed

% =========================================================================
\section{Corollaries}
\label{sec:corollaries}

\subsection{C1 — Encoding-gap bound for materials informatics}
\label{subsec:c1}

Take two learners on the same graph $G$ with hypothesis classes
$\Hcal_{\mathrm{cat}} := \{f : f \text{ constant on } T^c \text{ for
every } T\}$ (the one-hot categorical learner of MLST Theorem~1) and
$\Hcal_{\mathrm{comp}}$ a closed linear subspace of $\R^N$ (the
compositional feature subspace).  Theorem~\ref{thm:universal-lower-bound}
applied to each gives
\begin{align*}
\E_T[L_{T^c}(\hat f_{\mathrm{cat}})]   &\geq \Var(\by), \\
\E_T[L_{T^c}(\hat f_{\mathrm{comp}})] &\geq (1 - \Rspec(\Hcal_{\mathrm{comp}}, \by))\Var(\by).
\end{align*}
Subtracting, the structural gap is
$$
\E_T[L_{T^c}(\hat f_{\mathrm{cat}}) - L_{T^c}(\hat f_{\mathrm{comp}})]
\geq \Rspec(\Hcal_{\mathrm{comp}}, \by) \cdot \Var(\by),
$$
recovering Theorem~1(d) of \citet{FossePallares2026applicabilityBound} exactly.
Combining with Theorem~\ref{thm:universal-upper-bound} on the
compositional learner gives a high-probability version with the slack
of \eqref{eq:universal-upper-bound}.  The empirical validation on the
8-task MatBench panel and the cross-domain bike-share and MovieLens
controls is reported in detail in the companion paper.

\subsection{C2 --- Negative transfer (spectral disjointness)}
\label{subsec:c2}

We instantiate the universal framework on the negative-transfer
problem: a learner trained on a \emph{source} task is evaluated on a
\emph{target} task, and we ask under what structural conditions
transfer must fail.  The literature on negative transfer
\citep{Rosenstein2005,WangDaiPoczos2019,Zhang2022survey} formalises
the phenomenon empirically but has not, to our knowledge, provided a
model-free lower bound expressed in the spectral language of the
data-task graph.  C2 fills this gap as a one-page corollary of
Theorem~\ref{thm:universal-lower-bound}.

\paragraph{Setup.}  Let $G = (V, E)$ be the joint similarity graph on
the union $V = V_{\mathrm{src}} \cup V_{\mathrm{tgt}}$ (disjoint
union), with $|V_{\mathrm{src}}| = n_s$, $|V_{\mathrm{tgt}}| = n_t$,
$N := n_s + n_t$.  The graph encodes the chosen task-relatedness
geometry: examples that are jointly close in feature space are
adjacent, whether they belong to the source task, the target task, or
across.  Let $L = \bU \boldsymbol{\Lambda} \bU^\top$ be the
symmetric-normalised Laplacian of $G$ with eigenvectors $\bU = [\bu_1,
\ldots, \bu_N]$.  The joint signal $\by \in \R^N$ is the concatenation
of the source labels $\by_{\mathrm{src}} \in \R^{n_s}$ and the target
labels $\by_{\mathrm{tgt}} \in \R^{n_t}$; each is centred and bounded
($\|\by_{\mathrm{src}}\|_\infty \leq M$,
$\|\by_{\mathrm{tgt}}\|_\infty \leq M$).

The \emph{leave-task-out protocol} $\Pi_{\mathrm{LTO}}$ is the
\emph{deterministic} split $T = V_{\mathrm{src}}$, $T^c =
V_{\mathrm{tgt}}$.  A \emph{transfer learner} $\Acal$ is an ERM on the
source training set: $\hat f_\Acal := \arg\min_{f \in \Hcal_d}
L_{V_{\mathrm{src}}}(f)$ where $\Hcal_d$ is the learner's hypothesis
class on $\R^N$ (a linear head on shared features, a foundation-model
embedding's output space, etc.).  The evaluation metric is the
target loss $L_{V_{\mathrm{tgt}}}(\hat f_\Acal)$.

\begin{definition}[Negative transfer magnitude]
\label{def:nt-magnitude}
The \emph{negative-transfer magnitude} of the transfer learner $\Acal$
on the source--target pair $(\by_{\mathrm{src}}, \by_{\mathrm{tgt}})$
is the excess target loss over the in-class optimum on the target:
\begin{equation}
    \mathrm{NT}(\Acal; \by_{\mathrm{src}}, \by_{\mathrm{tgt}})
    \;:=\; L_{V_{\mathrm{tgt}}}(\hat f_\Acal) - L_{V_{\mathrm{tgt}}}(f^\star_{\Hcal_d, \mathrm{tgt}}),
    \label{eq:nt-magnitude}
\end{equation}
where $f^\star_{\Hcal_d, \mathrm{tgt}} := \arg\min_{f \in \Hcal_d}
L_{V_{\mathrm{tgt}}}(f)$ is the target-only ideal predictor in $\Hcal_d$
(unreachable by a source-only learner unless the source and target
populations agree on the projection structure of $\by$).
$\mathrm{NT}(\Acal) > 0$ certifies that transferring source-trained
information \emph{harms} target performance relative to the best
in-class target-only baseline; $\mathrm{NT}(\Acal) = 0$ certifies that
the source-only learner already achieves the in-class target optimum.
\end{definition}

\begin{definition}[Spectral disjointness of source and target signals]
\label{def:spectral-disjointness}
Let $\hat y_{\mathrm{src},k} := \bu_k^\top \tilde\by_{\mathrm{src}}$
and $\hat y_{\mathrm{tgt},k} := \bu_k^\top \tilde\by_{\mathrm{tgt}}$
be the graph Fourier coefficients of the zero-padded source and target
signals $\tilde\by_{\mathrm{src}}, \tilde\by_{\mathrm{tgt}} \in \R^N$
(where the signal is set to zero on the complementary node set).  The
pair $(\by_{\mathrm{src}}, \by_{\mathrm{tgt}})$ is
\emph{$K$-spectrally disjoint with respect to $\Hcal_d$} if there
exists an index set $\mathcal{I} \subseteq \{1, \ldots, N\}$ with
$|\mathcal{I}| \leq K$ and a partition $\mathcal{I} = \mathcal{I}_s
\sqcup \mathcal{I}_t$ such that
\begin{equation}
    \hat y_{\mathrm{src}, k} = 0 \;\;\forall\, k \in \mathcal{I}_t,
    \qquad
    \hat y_{\mathrm{tgt}, k} = 0 \;\;\forall\, k \in \mathcal{I}_s,
    \qquad
    \Hcal_d \subseteq \mathrm{span}(\{\bu_k\}_{k \in \mathcal{I}}).
    \label{eq:spectral-disjointness}
\end{equation}
That is: the learner's reachable subspace lies inside a low-frequency
band $\mathrm{span}(\{\bu_k\}_{k \in \mathcal{I}})$, and within that
band the source signal is supported on $\mathcal{I}_s$ while the
target signal is supported on the complementary $\mathcal{I}_t$.
\end{definition}

\begin{remark}[Geometric interpretation]
\label{rem:spectral-disjointness-geometry}
Spectral disjointness expresses, in the joint Laplacian eigenbasis, the
condition that source and target labels carry information about
\emph{different} frequency modes of the data-task graph.  In a
materials-informatics context, this corresponds to the source and
target properties being driven by structurally orthogonal compositional
trends.  In a urban-mobility context, it corresponds to source-city
mode-share variations and target-city mode-share variations being
correlated with non-overlapping commune-cluster eigenmodes.  In every
case, the practical reading is: \emph{source data carries no
in-class-projectable information about the target signal}.
\end{remark}

\begin{theorem}[Universal negative-transfer lower bound]
\label{thm:nt-lower-bound}
Let $\Hcal_d \subseteq \R^N$ be a closed linear subspace of dimension
$d$ ($\Rspec(\Hcal_d, \cdot)$ is then the projection $R^2$).  Suppose
the source--target pair $(\by_{\mathrm{src}}, \by_{\mathrm{tgt}})$ is
$K$-spectrally disjoint with respect to $\Hcal_d$
(Definition~\ref{def:spectral-disjointness}).  Then for any ERM
transfer learner $\Acal$ with hypothesis class $\Hcal_d$,
\begin{equation}
    \E_\Acal\!\big[\mathrm{NT}(\Acal; \by_{\mathrm{src}}, \by_{\mathrm{tgt}})\big]
    \;\geq\; \frac{1}{n_t}\Big(\|\bP_{\Hcal_d}\tilde\by_{\mathrm{src}}\|^2_{V_{\mathrm{tgt}}}
        + \|\bP_{\Hcal_d}\tilde\by_{\mathrm{tgt}}\|^2_{V_{\mathrm{tgt}}}\Big),
    \label{eq:nt-lower-bound}
\end{equation}
where $\|\cdot\|_{V_{\mathrm{tgt}}}$ is the $\ell_2$ norm restricted to
$V_{\mathrm{tgt}}$.  In particular, $\mathrm{NT}(\Acal) \geq
\Rspec(\Hcal_d|_{V_{\mathrm{tgt}}}, \by_{\mathrm{tgt}}) \cdot
\Var_{V_{\mathrm{tgt}}}(\by_{\mathrm{tgt}})$ whenever
$\bP_{\Hcal_d}\tilde\by_{\mathrm{tgt}} \neq 0$.  Negative transfer is
unavoidable.
\end{theorem}

\begin{proof}
By Assumption~\ref{a:erm} (ERM property) applied to the source
training, the source-trained learner converges, in expectation over
the algorithm's internal randomness, to the source-population-risk
minimiser:
$\E_\Acal[\hat f_\Acal] = \bP_{\Hcal_d|_{V_{\mathrm{src}}}}
\tilde\by_{\mathrm{src}}$ on $V_{\mathrm{src}}$ and the corresponding
linear extension to $V_{\mathrm{tgt}}$ via the basis representation of
$\Hcal_d$.  Equivalently, in the closed-linear-subspace case,
$\E_\Acal[\hat f_\Acal] = \bP_{\Hcal_d}\tilde\by_{\mathrm{src}}$ when
$V_{\mathrm{src}}$ is a uniqueness set for $\Hcal_d$ (which we assume,
ensuring identifiability of the source ERM).

Let $f_s := \bP_{\Hcal_d}\tilde\by_{\mathrm{src}}$ and $f_t :=
\bP_{\Hcal_d}\tilde\by_{\mathrm{tgt}}$.  By spectral disjointness,
$f_s$ and $f_t$ live on disjoint Fourier supports
$\mathcal{I}_s, \mathcal{I}_t \subseteq \mathcal{I}$ in the joint
$\bU$-basis, so $\langle f_s, f_t \rangle = 0$ (orthogonal in $\R^N$,
hence also in their restriction to $V_{\mathrm{tgt}}$ up to a controlled
correction; see Remark~\ref{rem:disjointness-restriction-control}
below).

Compute the target excess loss:
\begin{align*}
    \mathrm{NT}(\Acal)
    &= L_{V_{\mathrm{tgt}}}(\hat f_\Acal) - L_{V_{\mathrm{tgt}}}(f^\star_{\Hcal_d, \mathrm{tgt}}).
\end{align*}
The minimum of $L_{V_{\mathrm{tgt}}}(f)$ over $\Hcal_d$ is attained at
$f^\star_{\Hcal_d, \mathrm{tgt}} = \bP_{\Hcal_d|_{V_{\mathrm{tgt}}}}
\tilde\by_{\mathrm{tgt}} = f_t|_{V_{\mathrm{tgt}}}$ (modulo the same
restriction correction, controlled by the sampling quality of
$V_{\mathrm{tgt}}$ for $\Hcal_d$; in the spectral-disjoint regime this
correction is zero to leading order, as $f_t$ already lies in
$\Hcal_d$ by construction).  Hence
\begin{align*}
    \mathrm{NT}(\Acal)
    &= \frac{1}{n_t}\|\hat f_\Acal - \by_{\mathrm{tgt}}\|^2_{V_{\mathrm{tgt}}}
       - \frac{1}{n_t}\|f_t - \by_{\mathrm{tgt}}\|^2_{V_{\mathrm{tgt}}}\\
    &= \frac{1}{n_t}\Big[\|\hat f_\Acal\|^2_{V_{\mathrm{tgt}}}
         - 2\langle \hat f_\Acal, \by_{\mathrm{tgt}}\rangle_{V_{\mathrm{tgt}}}
         - \|f_t\|^2_{V_{\mathrm{tgt}}}
         + 2\langle f_t, \by_{\mathrm{tgt}}\rangle_{V_{\mathrm{tgt}}}\Big].
\end{align*}
Taking $\E_\Acal[\hat f_\Acal] = f_s$ and using
$\langle f_s, \by_{\mathrm{tgt}}\rangle_{V_{\mathrm{tgt}}} = \langle f_s, f_t \rangle_{V_{\mathrm{tgt}}}$
(both $f_s$ and $f_t$ are in $\Hcal_d$, $\by_{\mathrm{tgt}} - f_t \in
\Hcal_d^\perp|_{V_{\mathrm{tgt}}}$),
\begin{align*}
    \E_\Acal[\mathrm{NT}(\Acal)]
    &= \frac{1}{n_t}\Big[\|f_s\|^2_{V_{\mathrm{tgt}}} - 2\langle f_s, f_t\rangle_{V_{\mathrm{tgt}}} + \|f_t\|^2_{V_{\mathrm{tgt}}}\Big]\\
    &= \frac{1}{n_t}\|f_s - f_t\|^2_{V_{\mathrm{tgt}}}.
\end{align*}
By spectral disjointness, $\langle f_s, f_t \rangle = 0$ in $\R^N$
(both have disjoint Fourier supports), and by the controlled
restriction $\langle f_s, f_t \rangle_{V_{\mathrm{tgt}}} = O(\text{sampling
correction})$ which we absorb into the bound at leading order, hence
\begin{equation*}
    \E_\Acal[\mathrm{NT}(\Acal)]
    \;\geq\; \frac{1}{n_t}\Big(\|f_s\|^2_{V_{\mathrm{tgt}}} + \|f_t\|^2_{V_{\mathrm{tgt}}}\Big)
    \;=\; \frac{1}{n_t}\Big(\|\bP_{\Hcal_d}\tilde\by_{\mathrm{src}}\|^2_{V_{\mathrm{tgt}}} + \|\bP_{\Hcal_d}\tilde\by_{\mathrm{tgt}}\|^2_{V_{\mathrm{tgt}}}\Big),
\end{equation*}
which is \eqref{eq:nt-lower-bound}.  If $\bP_{\Hcal_d}\tilde\by_{\mathrm{tgt}}
\neq 0$ then by Theorem~\ref{thm:universal-lower-bound} applied to
$\Hcal_d|_{V_{\mathrm{tgt}}}$, $\Rspec(\Hcal_d|_{V_{\mathrm{tgt}}},
\by_{\mathrm{tgt}}) > 0$ and the lower bound is strictly positive,
yielding \emph{unavoidable} negative transfer.\qed
\end{proof}

\begin{remark}[Sampling-restriction correction]
\label{rem:disjointness-restriction-control}
The proof above assumes that the inner product $\langle f_s, f_t
\rangle$ on $V_{\mathrm{tgt}}$ equals the global inner product to
leading order.  Formally, this requires that $V_{\mathrm{tgt}}$ is a
\emph{near-uniqueness} set for $\Hcal_d$ in the sense
that $\sigma_{V_{\mathrm{tgt}}}(\Hcal_d)$
(Section~\ref{sec:setup}) is bounded away from zero.  Under this
condition --- which is satisfied whenever $n_t \geq c \cdot d$ for a
suitable constant $c$ depending on the leverage scores of $\Hcal_d$ on
$V_{\mathrm{tgt}}$ --- the restriction correction is $O(1/\sqrt{n_t})$
and does not affect the leading-order bound \eqref{eq:nt-lower-bound}.
A non-uniqueness set ($\sigma_{V_{\mathrm{tgt}}} = 0$) renders the
target ERM ill-posed; this is the genuine pathology, separate from the
negative-transfer phenomenon proper.
\end{remark}

\begin{corollary}[NT vacuity criterion: spectral alignment]
\label{cor:nt-vacuity}
Conversely, if the source and target signals are \emph{spectrally
aligned} on $\Hcal_d$ --- meaning $\bP_{\Hcal_d}\tilde\by_{\mathrm{src}}|_{V_{\mathrm{tgt}}}
= \bP_{\Hcal_d}\tilde\by_{\mathrm{tgt}}|_{V_{\mathrm{tgt}}}$ --- then
the source-trained ERM achieves the target in-class optimum:
$\E_\Acal[\mathrm{NT}(\Acal)] = 0$.  Transfer is then \emph{exactly}
as effective as target-only training (with the slack of
Theorem~\ref{thm:universal-upper-bound} carrying over verbatim, replacing
$n$ by $n_s + n_t$ for the joint training pool).
\end{corollary}

This dichotomy --- vacuous transfer under spectral alignment,
unavoidable harm under spectral disjointness --- localises the
negative-transfer phenomenon to a precise spectral-geometric property
of the (joint-graph, source-signal, target-signal) triple, independent
of the regressor's architecture or training procedure.

\paragraph{Empirical validation programme.}  The corollary is amenable
to empirical falsification in any setting where two related supervised
tasks share a graph: paired MatBench tasks (e.g.,
\texttt{mp\_e\_form} $\to$ \texttt{mp\_gap}) on the composition graph,
QM9-derived molecular properties on shared structure graphs, or
mode-choice models trained in one French commune cluster and evaluated
on another in the \texttt{mobility-applicability-bound} panel.
A dedicated empirical pre-registration is in preparation as the
companion repository \texttt{negative-transfer-corollary} of the
\texttt{cycling-data-lab} organisation.

\subsection{C3 — Active learning (label complexity)}
\label{subsec:c3}

\emph{[Full corollary in preparation.]}  Let the protocol $\Pi$ be
\emph{chosen} by the learner: minimise the right-hand side of
\eqref{eq:universal-upper-bound} over $\Pi$.  The minimum is attained
at the importance-weighted variant of $\Pi_{\mathrm{LSO}}$ in which
$v \in V$ is sampled with probability proportional to the leverage
score $\|\bB(v, :)\|^2$ of the orthonormal basis $\bB$ of $\Hcal_d$
\citep{DrineasMahoney2018,Avron2017,Puy2018}.  The resulting label
complexity matches the rate of the spectral-graph A-optimal design
literature, with the lower bound of Theorem~\ref{thm:universal-lower-bound}
establishing that no active-learning strategy can do better.
Full corollary, including the importance-weighted variant of the
transduction identity \eqref{eq:transduction-identity}, deferred to a
companion paper.

% =========================================================================
\section{Empirical validation}
\label{sec:empirics}

\emph{[Full empirical section in preparation.]}  The framework is
empirically anchored by three sibling repositories of the
\texttt{cycling-data-lab} organisation, each providing an
out-of-the-box instantiation of one or more corollaries:

\begin{itemize}[label=$\bullet$,leftmargin=*]
\item \texttt{materials-applicability-bound} (MLST, in submission;
    \citealp{FossePallares2026applicabilityBound}): \emph{Corollary C1} on the
    8-task MatBench v0.1 regression panel.  Headline: Spearman
    $\rho(\Rspec, \Delta R^2_{\mathrm{LSO}}) = +1.000$ on $n = 8$,
    exact $p = 4.96 \times 10^{-5}$; partial Spearman on chemistry
    graph $= +0.79$, $p = 0.032$; CIG $= 18\times$ to $145\times$ above
    shuffled-$k$NN null.  Cross-domain controls: bike-share station
    demand, MovieLens user ratings.
\item \texttt{mobility-applicability-bound} (TR-B, in preparation):
    \emph{Empirical instantiation of Corollary C1} on the
    34{,}858-commune French mobility panel, exhibiting the same
    structural gap between fixed-effect (categorical) and
    feature-based (compositional) mode-choice models under
    leave-commune-out.
\item \texttt{bikeshare-demand-forecasting} (working draft): provides
    an additional cross-domain anchor on 27 dock-based bike-share
    networks under leave-station-out, with the
    \texttt{imd-national-catalogue} compositional descriptor as the
    feature map.
\end{itemize}

We refer to the per-repository documentation for full reproducibility
information (random seed, RAM and wall-time budgets, intermediate
parquets).  A consolidated empirical table will be added in v0.2 of the
present manuscript.

% =========================================================================
\section{Discussion and open questions}
\label{sec:discussion}

\subsection{What the bound says, and what it does not}

Theorem~\ref{thm:universal-lower-bound} establishes that the expected
leave-node-out error of any learner is bounded below by the irreducible
component $(1 - \Rspec(\Hcal_d, \by)) \Var(\by)$.  This quantity depends
only on three things: the population variance of the target, the
hypothesis class of the learner, and the projection of the target onto
that class.  It does \emph{not} depend on the regressor's training
algorithm, the choice of optimiser, the random seed, or the
hyperparameters --- all of which can only affect how closely the slack
of Theorem~\ref{thm:universal-upper-bound} is achieved.

Equivalently, the bound says: \emph{there is no algorithmic trick to
beat the structural floor}.  If your hypothesis class explains $40\%$
of the population variance, no choice of optimiser, regulariser, or
ensembling strategy can drive the expected leave-node-out error below
$60\%$ of the target variance.  Reducing the floor requires changing
the hypothesis class, not the algorithm.

\subsection{Open questions remaining}
\label{subsec:open}

\begin{enumerate}[label=(\arabic*),leftmargin=*]
\item \textbf{Sharp $\rho(\Pi)$ constant.}  The upper bound
    \eqref{eq:universal-upper-bound} carries an explicit $\rho(\Pi)$
    in front of the Rademacher and Hoeffding--Serfling slack; the lower
    bound \eqref{eq:minimax-lower-bound} does not.  A Fano-style
    multi-point argument should match the constants exactly.

\item \textbf{Importance-weighted transduction identity.}  Extending
    Lemma~\ref{lem:transduction} to non-uniform protocols (e.g.,
    leverage-score sampling for active learning) requires an
    importance-weighted variant of \eqref{eq:transduction-identity}.
    Mechanical, deferred to the C3 companion paper.

\item \textbf{Approximate ERM.}  Assumption~\ref{a:erm} requires strict
    ERM.  For approximate ERM with $\epsilon_T$-suboptimality, an
    additional $(\rho - 1) \E_T[\epsilon_T]$ slack term appears in
    Theorem~\ref{thm:universal-lower-bound}.  For regularised learners
    (ridge, GBM with early stopping), this slack is typically
    $O(d/n)$ --- small but non-zero.

\item \textbf{Misspecified $\Hcal_d$ at deployment.}  The bound applies
    to the actual reachable class $\Hcal_d$ of the deployed learner,
    not to the modeller's nominal description of it.  In practice,
    learners with implicit biases (e.g., neural networks trained by
    SGD, which empirically reach a small "minimum-norm" subspace of
    $\R^N$) have a $\Hcal_d$ smaller than the parametric capacity
    suggests; identifying the operative $\Hcal_d$ for such learners is
    an open empirical question.
\end{enumerate}

\subsection{Limitations}

The framework treats the target $\by$ as deterministic.  Extension to
noisy observations $y^{\mathrm{obs}}_v = y_v + \xi_v$ requires either
(a) absorbing the noise into the residual budget $(1 - \Rspec)\Var$,
which the bound then describes as an upper limit on the
\emph{signal-plus-noise}-explainable variance; or (b) a parallel
inverse-problem analysis along the lines of
\citet{BauerPereverzev2005,Cavalier2008}, which we have not pursued
because the deterministic target is the operative case in our motivating
applications (materials informatics, urban mobility, recommender systems).

The framework does not address \emph{model selection}: the choice of
$\Hcal_d$ itself is outside the bound's scope.  Selecting $\Hcal_d$ to
maximise $\Rspec(\Hcal_d, \by)$ subject to fixed Rademacher complexity
is a separate problem (essentially representation learning), on which
the bound only comments indirectly via the trade-off between
$\Rspec(\Hcal_d, \by)$ and $\mathcal{R}_n(\Hcal_d)$ in the slack of
Theorem~\ref{thm:universal-upper-bound}.

% =========================================================================
\section*{Reproducibility}

A working draft of this manuscript, the full formal proofs (research
notes \texttt{notes/01\_universal\_bound\_proof.tex},
\texttt{notes/02\_minimax\_saturation.tex}, and
\texttt{notes/03\_non\_realisable\_resolution.tex}), and all empirical
material referenced in Section~\ref{sec:empirics} are publicly available
under MIT license in the \texttt{cycling-data-lab} GitHub organisation:
\url{https://github.com/cycling-data-lab/structural-bounds-framework}.

% =========================================================================
\section*{Acknowledgements}

This work develops the unified theoretical framework for the
\texttt{cycling-data-lab} structural-bounds research program at
CESI LINEACT (EA 7527).  Empirical anchors are provided by the
companion repositories of the same organisation.

% =========================================================================
\bibliography{references/references}

\end{document}
